---
title: "Functional Genomics Part 3"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r}
library(here)
library(tidyverse)
library(markerGeneProfile)
```

# Estimating inhibitory SST neuron proportions across the adult cerebral cortex using the Allen Human Brain Atlas using the aparc Freesurfer space (Desikan–Killiany cortical atlas)

### Read and process the marker genes as before

```{r}
# read marker genes
acc_markers = read_csv(here('marker_genes','ACC_results.csv'))

cell_types = acc_markers$cluster %>% unique()
```

```{r}
marker_list <- lapply(cell_types, function(cell_type){
  print(cell_type)
  return(acc_markers %>% filter(cluster == cell_type) %>% pull(gene) %>% unlist())
})
names(marker_list) <- cell_types

```

### read Allen Brain Atlas data in the freesurfer space from figshare and fix first columns
### https://figshare.com/articles/A_FreeSurfer_view_of_the_cortical_transcriptome_generated_from_the_Allen_Human_Brain_Atlas/1439749

```{r}
## reads the data from figshare if it has not been already downloaded

if (file.exists(here("allen_brain", "expression.tsv"))) {
  expression <- read_tsv(here("allen_brain", "expression.tsv"))
  expression_region_metadata <- read_tsv(here("allen_brain", "expression_metadata.tsv"))
} else {
  expression <- read_tsv("https://ndownloader.figshare.com/files/7013300")
  expression_region_metadata <- read_tsv("https://ndownloader.figshare.com/files/3342527")
}

## change one of the column names
expression %<>% dplyr::rename(gene_symbol = X1)

## fix one of the column names
expression_region_metadata %<>% dplyr::rename(region_name = X1)
```

### Filter out genes with low expression

```{r}

expressionMatrix <- expression %>% select( Probe = gene_symbol, Gene.Symbol = gene_symbol,everything())

medExp = expressionMatrix %>%
  sepExpr() %>% {.[[2]]} %>%
  unlist %>% median
expressionMatrix = mostVariable(expressionMatrix, threshold = medExp, threshFun=median)

```

### run MGP to get estimates

```{r}
estimations =  mgpEstimate(exprData=expressionMatrix,
                           genes=marker_list,
                           geneColName='Gene.Symbol',
                           outlierSampleRemove=F, # should outlier samples removed. This is done using boxplot stats.
                           geneTransform =NULL, #markers are human genes
                           groups=NULL, #if there are experimental groups provide them here. if not desired set to NULL
                           seekConsensus = FALSE, # ensures gene rotations are positive in both of the groups
                           removeMinority = TRUE) # removes genes if they are the minority in terms of rotation sign from estimation process
```

```{r}
mgp_estimates <- as.data.frame(estimations$estimates) %>% as_tibble(rownames = "region_name")
mgp_estimates %<>% select(region_name, Inh_SST)
colnames(mgp_estimates)
mgp_df <- inner_join(expression_region_metadata, mgp_estimates, by = "region_name")

```

### Inspect SST proportions - low in occipital regions and high in frontal regions
```{r}
mgp_df %>% arrange(Inh_SST) %>% print() #acending
mgp_df %>% arrange(-Inh_SST) %>% print() #descending
```

### Write out for future analyses

```{r}
dir.create(here("results", "Allen_SST_estimates"), recursive = T)
mgp_df %>% write_csv(here("results", "Allen_SST_estimates", "Regional_SST_estimates.csv"))
```

## Optional Exercises

* Visualize SST estimates using tksurfer from FreeSurfer (https://figshare.com/articles/A_FreeSurfer_view_of_the_cortical_transcriptome_generated_from_the_Allen_Human_Brain_Atlas/1439749)
